---
title: "Competition indices with forestCI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Competition indices with forestCI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 5, dpi = 150)
library(forestCI)
library(data.table)
```

## The one object everything reads from

A competition index is a summary of a neighbourhood, so the first job is saying
what a neighbourhood is. `as_stand()` does that once: it names the columns,
fixes the plot geometry, attaches species traits, and derives basal area, crown
widths and crown ratio. Every other function in the package takes the result and
nothing else.

```{r}
st <- pef_stand()
st
```

These are two stem-mapped permanent plots from the Penobscot Experimental
Forest, each 0.0809 ha. Every tree carries a distance and bearing from plot
centre, so the stand is spatial and the distance-dependent family is available.

```{r stemmap, fig.cap = "Figure 1. Stem maps of the two example plots, symbol area proportional to basal area."}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1), bty = "l")
for (p in unique(st$trees$plot_id)) {
  d <- st$trees[st$trees$plot_id == p, ]
  plot(d$x, d$y, cex = d$dbh / 12, pch = 21,
       bg = ifelse(d$sp_type == "SW", "#2b6a4d", "#b07a2a"),
       asp = 1, xlab = "Easting (m)", ylab = "Northing (m)", main = p)
  symbols(0, 0, circles = 16.05, inches = FALSE, add = TRUE, fg = "grey60")
}
par(op)
```

## Without coordinates

Most inventories have no stem map. That is not a failure case here. Drop the
coordinates and the distance-independent family still computes, the rest is
reported as skipped, and the returned table keeps the same shape.

```{r}
flat <- as.data.frame(st$trees)
st_flat <- as_stand(flat, plot = "plot_id", tree = "tree_id",
                    species = "species", dbh = "dbh", expf = "expf",
                    quiet = TRUE)
ci_flat <- competition_indices(st_flat)
head(ci_flat[, .(tree_id, dbh, ba, bal, ccf, ccfl, sdi, rd)])
```

## With coordinates

```{r}
ci <- competition_indices(st, radius = 6, edge = "exclude",
                          n_depth = 12L, n_azimuth = 36L, quiet = TRUE)
head(ci[, .(tree_id, species, dbh, bal, ccfl, hegyi, local_bal, apa, csax_rel)])
```

`edge = "exclude"` blanks the distance-dependent columns for trees the exterior
sector rule flags as plot edge trees. The distance-independent columns are left
alone, because they were never biased by the boundary in the first place.

## How much does spatial detail buy

The question the original work asked. Correlate each index against the others
and the answer is visible without fitting anything: the one-sided indices,
spatial or not, carry nearly the same information.

```{r corr, fig.cap = "Figure 2. Correlation among competition indices on the example plots. Spatial and non-spatial one-sided indices are close to collinear."}
keep <- c("bal", "ccfl", "local_bal", "hegyi", "martin_ek", "apa", "csax_rel")
m <- stats::cor(as.matrix(ci[, ..keep]), use = "pairwise.complete.obs")
op <- par(mar = c(6, 6, 2, 2))
image(seq_len(ncol(m)), seq_len(ncol(m)), t(m[nrow(m):1, ]), axes = FALSE,
      xlab = "", ylab = "", zlim = c(-1, 1),
      col = grDevices::hcl.colors(21, "Blue-Red 3", rev = TRUE))
axis(1, seq_len(ncol(m)), colnames(m), las = 2)
axis(2, seq_len(ncol(m)), rev(rownames(m)), las = 2)
box()
par(op)
```

## Crown profiles

The crown based indices are the reason the package carries a crown model at
all. One object controls the geometry everywhere.

```{r profiles, fig.cap = "Figure 3. Crown profiles of a 20 m red spruce with a 10 m crown under three profile families."}
h <- 20; hcb <- 10; lcw <- 4
zz <- seq(hcb, h, length.out = 200)
fams <- c("dual_exponent", "variable_exponent", "geometric")
cols <- c("#2b6a4d", "#b07a2a", "#4a4a4a")
plot(NA, xlim = c(-2.2, 2.2), ylim = c(hcb - 0.5, h + 0.5),
     xlab = "Crown radius (m)", ylab = "Height above ground (m)", bty = "l")
for (k in seq_along(fams)) {
  p <- crown_profile(fams[k])
  r <- crown_radius_at(p, zz, lcw, h, hcb, widest = 0.8,
                       up_exp = 3, lo_exp = 3, shape = "g", cr = 0.5, hw = 0)
  lines(r, zz, col = cols[k], lwd = 2)
  lines(-r, zz, col = cols[k], lwd = 2)
}
legend("topright", fams, col = cols, lwd = 2, bty = "n")
```

`crown_profile_info()` states the provenance of the constants, which matters
because the `variable_exponent` coefficients shipped here are a calibration to
the package's own defaults rather than a field fit.

```{r}
crown_profile_info(crown_profile("variable_exponent"))[, c("constant", "value")]
```

## Adding species

```{r}
tr <- species_traits(data.frame(
  code = c("DF", "WH"), sp_type = "SW",
  sg = c(0.45, 0.42), shade = c(2.78, 4.83),
  widest = 0.75, up_exp = 3, lo_exp = 3, shape = "g"))
tr[tr$code %in% c("DF", "WH"), c("code", "sp_type", "sg", "shade", "shape")]
```

Only `code` and `sp_type` are required. Anything left out takes the softwood or
hardwood default, which is also what an unrecognised code falls back to, with a
message rather than silently.
